**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-08-23 submission/manuscript.Rmd
add interpretation data and fix after mmeting 4

Third, fitting of these models require selecting appropriate hyperparameters. Hyperparameters are the rules that are learned from the training set in a classification algorithm. For example, in the linear models the regularization term (C) is a hyperparameter that indicates the penalty for overfitting. Similar to regularization term C, all hyperparameters are tuned to find the best model. We selected hyperparameters by performing 100 five-fold cross-validation (CV) repeats on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the best hyperparameter values for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model’s predictions are no different from random chance.

@@

@@

We interpreted the feature importance in the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median absolute feature weights for each OTU in the models [Figure 3]. The OTUs that had the largest median rank and drove the detection of SRNs across the linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 1239, OTU 742), Roseburia (OTU 822, OTU00768), Clostridium (OTU 609), Blautia (OTU 659, OTU 12), Faecalibacterium (OTU 15), Gamella (OTU 426) and Peptostreptococcus (OTU 367) [Figure 3]. Among the five highest ranked OTUs across these three models, it was encouraging that OTU 1239 was shared among each of the models and that 4 OTUs were shared among the five highest ranked OTUs for both of the L2-regularized models.

We interpreted the feature importance in the non-linear models using permutation importance. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest. The OTUs that had the largest impact and drove the detection of SRNs across the non-linear models contained sequences that were most similar to members of the Lachnospiraceae (OTU 50), Ruminococcaceae (OTU 958, 8, 19), Flavonifractor (OTU 629), Howardella (OTU 299), Peptostrepococcus (OTU 367), Clostridium (OTU 58), Anaerostipes (OTU 10), Porphyromonas (OTU 110), Coprobacillus (OTU 477), Parvimonas (OTU 361), Roseburia (OTU 822) [Figure 4]. Among the twenty OTUs with the largest impact across these four models, there was only one OTU (OTU 822) that was shared among all of the models; however, among the tree-based models, we found three OTUs (OTU 367, OTU 58 and OTU 110) to be consistently important. Similarly, in the random forest and XGBoost models, four of the twenty most important OTUs were shared between the models.

When we analyzed the L1-regularized SVM with linear kernel model using permutation importance, the approach picked out many of the same OTUs (e.g. OTU00822, OTU01239, OTU00609) that were important when we interpreted the model using feature rankings based on weights [Figure 3]. Similarly, L2-regularized SVM and L2-regularized logistic regression  each identified many OTUs that were selected by both interpretation methods.

Interestingly, for the all the models, except the poorly performing L1-regularized SVM and decision tree, the variation in the impact of these important OTUs was minimal. This supports the findings of others that a classification requires numerous OTUs, rather the presence of a single or a few of the OTUs [@sze_leveraging_2018].

__The computational efficiency of each ML model.__

We compared the training times of the seven ML models. As expected, the training times increased as the complexity of a ML model and the number of tuned hyperparameter settings increased and the linear models trained faster than non-linear models [Figures S1-S2; Figure 5]. When we subsetted the size of the training dataset, we observed a [linear, quadratic, etc] relationship between the size of the dataset and the training time for the random forest model.

@@

Overview
•	Establish pipeline that can be generalized to any modeling method
•	Linear/non-linear not a bfd
•	Interpretability tradeoff
Point 1
•	Importance for transparency, picking the most complex model is not necessarilty the best model
•	Good / better / best:  Table S1
•	“best” has multiple meanings
Point 2
•	Our paper is concerned with developing models to gain greater biological insights into what’s going on
•	Still need biological validation and further experimentation
Point 3
•	These models show great opportunity to use microbiome to make diagnoses
•	We talked about classification, but others have used it in microbiome data to do regression, mulit-class – pipeline is the same
•	Others have used this on shotgun metagenomic data and collinearity is expected to be an even larger problem since 5000 genes are all correlated with each other
•	Deployment is a different question that requires independent datasets to test on; we currently lack decent sized parallel datasets where the same type of sequence data and clinical data have been collected
•	We also have a problem with OTU-based approaches
Conclusion/Future direction
•	ML is awesome and there’s always new methods; our framework gives structure to investigators wanting to try these out. We encourage reseraches to apply multiple models to their data and report the results
•	Things like neural nets are attractive and hold great promise, but we need a ton more data to apply and they are impossible to interpret

Microbiome studies use ML models, often with a classification task, to predict a disease but also to learn which microbes are indicators of that disease [@zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018]. Achieving either of these tasks have far-reaching impact on human health, however ML as a tool in microbiome studies is still at its infancy. A framework is needed to develop rigorous ML models, to identify and overcome potential pitfalls. Previous studies generated workflows to allow ML to be widely used by the microbiome researchers [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. This study sets-up standards for developing and evaluating rigorous ML models for microbiome data [Table 1].

We benchmarked seven ML models with different classification algorithms to show that a clearly defined ML problem that is based on the goal of the microbiome study should inform our model selection. Our results showed that if the goal of a study is to learn the ecology behind a disease and to identify microbial biomarkers, we can create ML models that are inherently interpretable and easily trained without losing predictive power. In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models. However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest was a more complex ML model and could only be explained using methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking absolute feature weights of the trained model.

Even with interpretable models such as L2-regularized logistic regression, there are potential pitfalls when it comes to identifying biomarkers of a disease. As domain experts, we know that human-associated microbial communities have complex correlation structures that creates collinearity in the dataset. Collinearity is a severe problem and needs to be addressed to reliably interpret ML models [@dormann_collinearity:_2013]. In this study we used two different methods to interpret our linear models; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance. We observed differences in the OTU rankings between the two interpretation methods due to collinearity in the dataset. To avoid misinterpreting the models, once we identify the most important microbes, we should check for their relationships with other microbes as well. These relationships will help us generate new hypotheses about the ecology of the disease. These hypotheses needs to be tested with follow-up experiments to identify the true biomarkers of a disease.

In this study, we also established a rigorous ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. First of all, we used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the difference between cross-validation and test performance is low, this suggest the models are not overfit and that they will perform similar with similar data. In all the models, the difference between median cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models are generalizable and can be used to test similar new data. Furthermore, we performed the initial 80%-20% random datasplit 100 times in our ML pipeline. Depending on how the data is split, there is the chance of being overoptimistic about the predictive performance of a model. We showed that there was variability in AUROC values between different random data-splits in each of the models we tested. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average between different data-splits. The randomization and resampling of the initial data-split to create a held-out test set is a crucial step in the ML pipeline to develop robust ML models and to report reliable performance metrics. Additionally, we performed a full grid search for hyperparameter settings when training our ML models. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default. And finally, we used the AUROC metric in our study to evaluate the predictive performance of the ML models. AUROC is always random at the value 0.5 and is a robust metric when a dataset is imbalanced.

We used a balanced CRC dataset to develop ML models with a binary classification task. We did not evaluate multicategory classification methods or regression analyses to predict non-binary outcomes. However, the principles highlighted throughout this study [Table 1] apply to all ML modeling tasks with microbiome data. The models we built were generalizable despite the high number of features microbiome datasets usually have. The generalization performance of ML models depends on sample size. The more complex the model, the more data it will need. Our dataset had 490 samples, however microbiome studies that have smaller sample sizes would benefit from using less complex models. Our analysis was limited to shallow learning methods and did not explore deep learning methods such as neural networks. Microbiome datasets often suffer from having high dimensionality but low sample sizes which makes deep learning models prone to overfitting. There are studies that address overcoming these challenges in biomedical datasets [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018], however sudies that estalish frameworks with microbiome data are lacking. This would be an interesting direction for future work in microbiome studies.

>>>>

Third, fitting of these models require selecting appropriate hyperparameters. Hyperparameters are the rules that are learned from the training set in a classification algorithm. For example, in the linear models the regularization term (C) is a hyperparameter that indicates the penalty for overfitting. Similar to regularization term C, all hyperparameters are tuned to find the best model. We selected hyperparameters by performing 100 five-fold cross-validation (CV) repeats on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the best hyperparameter values for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model’s predictions are no different from random chance. To select the best performing hyperparameter, we performed a full grid search for hyperparameter settings when training our models. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new dataset used to generate a model. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.

@@

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S4]. The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models. These results showed that because the microbiome data had many features (6920 OTUs), it was important to train the models using appropriate sample sizes to avoid problems with generalizability and over-optimism. Furthermore, it was encouraging that even for a small number of samples, the 95% confidence interval included the median AUROC values for the larger subsetted datasets.

@@

First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3]. We also reviewed at the signs of feature weights to determine whether an OTU is associated with classifying a subject as being healthy or having an SRN - negative sign indicated being healthy and positive sign indicated having SRN. Among the highest ranked OTUs across these three models, it was encouraging that many of the same OTUs (e.g. OTU 50, 426, 609, 822, 1239) were important across all of the models. The benefit of this approach is that the results of the analysis are based on the trained model parameters and provide information regarding the sign and magnitude of the impact of each OTU. However, this approach is only possible with linear models.

Second, to analyze non-linear models we interpreted the feature importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact for each of these models, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had a minimal impact on model performance. For example, if OTU 367 were permuted across the diagnoses from the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if we permuted the same OTU from the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(round(RF$imp, 3))`. Effectively, the complexity of the communities was more fully represented in the better performing models[@sze_leveraging_2018; @wirbel_meta-analysis_2019]. At least in this case, it was not possible to distinguish between health and disease using a single OTU. While permutation analysis allowed us to gauge the importance of an OTU, the analysis is post-hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU is associated with classifying a subject as being healthy or having an SRN.

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S3]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S3], 17 out of the 20 top OTUs (e.g. OTU 609, 822, 1239) were deemed important by both approaches. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two approaches. These results indicate that both approaches are consistent in selecting the most important OTUs.

__The computational efficiency of each ML model.__ We compared the training times of the seven ML models. As expected, the training times increased with the complexity of the model and the number of tuned hyperparameter settings. Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5]. When we subsetted the size of the training dataset, we observed a [linear, quadratic, etc] relationship between the size of the dataset and the training time for the random forest model.

@@

Overview
•	Establish pipeline that can be generalized to any modeling method
•	Linear/non-linear not a bfd
•	Interpretability tradeoff
Point 1
•	Importance for transparency, picking the most complex model is not necessarilty the best model
•	Good / better / best:  Table S1
•	“best” has multiple meanings
Point 2
•	Our paper is concerned with developing models to gain greater biological insights into what’s going on
•	Still need biological validation and further experimentation

Point 3
•	These models show great opportunity to use microbiome to make diagnoses
•	We talked about classification, but others have used it in microbiome data to do regression, mulit-class – pipeline is the same
•	Others have used this on shotgun metagenomic data and collinearity is expected to be an even larger problem since 5000 genes are all correlated with each other
•	Deployment is a different question that requires independent datasets to test on; we currently lack decent sized parallel datasets where the same type of sequence data and clinical data have been collected
•	We also have a problem with OTU-based approaches
Conclusion/Future direction
•	ML is awesome and there’s always new methods; our framework gives structure to investigators wanting to try these out. We encourage reseraches to apply multiple models to their data and report the results
•	Things like neural nets are attractive and hold great promise, but we need a ton more data to apply and they are impossible to interpret

There is a growing awareness that many human diseases and environmental processes are not driven by a single organism, but are the product of multiple bacterial populations. Whereas traditional statistical approaches are useful for identifying those cases where a single organism can drive a process, ML methods offer the ability to use many features of these communities to identify those cases. If it is possible to classify communities reliably, then ML methods also offer the ability to identify those populations that are responsible for the classification. Because application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models [Table 1].

To test this framework, we benchmarked seven ML models to show that a clearly defined ML problem that is based on the goal of the microbiome study should inform our model selection. Our results showed that if the goal of a study to identify microbial biomarkers of disease, we can create ML models that are inherently interpretable and easily trained without losing predictive power. In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models. However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest was a more complex ML model and could only be explained using post-hoc methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights.

Even with interpretable models such as L2-regularized logistic regression, there are potential pitfalls when it comes to identifying biomarkers of a disease. Human-associated microbial communities have complex correlation structures. This can hinder our ability to reliably interpret models [@dormann_collinearity:_2013]. In this study we used two different interpretation methods; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance. The feature weights of correlated OTUs are influenced by and dependent on one another. To avoid misinterpreting the models, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. To avoid misinterpreting models due to correlation structures of the communities when we used permutation importance, we grouped correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features to identify the true underlying factors when making a classification.

In this study, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome. We used a held-out test set and we performed the initial 80%-20% random datasplit to create the held-out test set 100 times in our ML pipeline. The models we built with this pipeline were generalizable despite the high number of features microbiome datasets usually have.

We used a CRC dataset to develop ML models with a binary classification task. We did not evaluate multicategory classification methods or regression analyses to predict non-binary outcomes. However, the principles highlighted throughout this study [Table 2] apply to all ML modeling tasks with microbiome data. Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks. Microbiome datasets often suffer from having high dimensionality but low sample sizes which makes deep learning models prone to overfitting.

There are studies that address overcoming these challenges in biomedical datasets [@geman_deep_2018; @kocheturov_massive_2019; @kim_improved_2018], however sudies that estalish frameworks with microbiome data are lacking. This would be an interesting direction for future work in microbiome studies.
